Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-14 text/back_end_KS.qmd
LA revisions

For Asian students, the Glendale and Pasadena (17% versus 14% for all prospects), East Los Angeles (11% versus 5% for all prospects) and Covina and West Covina (17% versus 11% for all prospects) Geomarkets contributed a disproportionate larger shares, whereas nearly all other Geomarkets contributed disproportionately smaller or nearly equal shares.

@@

Contributions of each Geomarket to the pool of prospects from the Los Angeles metropolitan area based on first-generation college student status are presented in @fig-la-rq2-midSAT-ses [THIS SENTENCE IS UGLY. MAKE IT MORE LIKE FIG X SHOWS THE CONTRIBUTIONS EACH GEOMARKET MAKES BY FIRST GENERATION STATUS GROUP]. Only 889 of the more than 8,000 total prospects (11%) in this middle PSAT score range (1190-1260) pool were first-generation college students whose parents did not attend college. \[I'M NOT SURE WHERE YOU ARE GETTING THESE NUMBERS FROM; THIS IS THE WRONG FIGURE I THINK. AND WHEN YOU REVISE, THE LEDES ARE RIVSERSIDE, EAST LA AND, SOUTH CENTRAL (ONLY 2% OF PROSPECTS BUT 10% OF NO-COLLEGE)\]. Only three of the 11 Geomarkets contributed substantially larger shares of these first-generation prospects: East Los Angeles (35%), South and South Central Los Angeles (47%), and Riverside, San Bernardino, and Ontario (15%). However, disproportionate contributions become more spread across Geomarkets for the nearly 1,436 of the more than 8,000 total prospects (18%) who were first-generation college students whose parents attended but did not complete their degree. Overall, six of the 11 Geomarkets contributed larger shares of these prospects: Hollywood & Wilshire (25%), East Los Angeles (26%), South and South Central Los Angeles (28%), Long Beach (25%), Covina and West Covina (22%), and Riverside, San Bernardino, and Ontario (22%). This leads to the remaining Geomarkets - San Fernando Valley- West (80%), San Fernando Valley- East (78%), Glendale and Pasadena (77%), West Los Angeles & West Beach (87%), South Bay (81%) - contributing disproportionately larger shares of not first-generation college students relative to the overall prospect pool (71%).

[I THINK THE WAY YOU WRITE UP THIS FIGURE SHOULD BE CHANGED MORE SUBSTANTIALLY; I ADD SOME COMMENTS BELOW] Lastly, @fig-la-rq2-midSAT-combo presents Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity in the Los Angeles metropolitan area. [TAKE A TIP FROM THE CRIT QUANT FOLKS; YOU DON'T ALWAYS NEED TO START WITH WHITE; REVISE THIS TEXT TO START WITH A PARAGRAPH ON HISPANIC AND A PARAGRAPH ON ASIAN -- OR VICE VERSA; THOSE TWO ARE THE LEAD; AND THEN CAN TALK ABOUT WHITE AND BLACK. ]Among all White prospects (n=2,511) whose contact information was purchased in this middle PSAT score range, approximately 3% were first-generation college students whose parents did not attend college, 11% were first-generation college students whose parents attended some college but did not complete their degree, and 86% were not first-generation college students. @fig-la-rq2-midSAT-combo shows all Geomarkets contribute a nearly proportionate share of prospects by first-generation college students (across both no college and some college parents) relative to the overall pool. The South and South Central Los Angeles Geomarket stands out as one exception contributing a relatively larger share of White first-generation college students whose parents did not attend college (17%) and whose parent attended some college but did not complete their degree (33%) \[DON'T MENTION THIS; THIS IS 6 STUDENTS!\]. Other Geomarket that contributed signifcantly larger shares of White first-generation prospects whose parents did not complete their degree include Long Beach (17%), and Riverside, San Bernardino, and Ontario (19%).

[THE LEAD IS EAST LA WHICH HAS 251 ASIAN STUDENTS AND IS WAY DIFFERENT THAN ANY OTHE IN FIRSTGEN; AND THEN WILSHIRE IS SORTA CLOSE] Geomarkts contributions to first-generation college students was mixed across other racial/ethnic categories in @fig-la-rq2-midSAT-combo. For Asian prospects (11% first-generation no college, 21% first-generation some college, and 68% not first-generation), the Hollywood & Wilshire (12%) and East Los Angeles (36%) Geomarkets contributed larger proportions of first-generation college students whose parents did not attend college. The San Fernando Valley- East (28%), Hollywood & Wilshire (35%), East Los Angeles (25%), South and South Central Los Angeles (44%) [ONLY 9 STUDENTS!], Long Beach (27%), and Covina and West Covina (25%) Geomarkets also contributed larger proportions of first-generation college students whose parents attended college but did not complete their their degree.

For Black prospects (5% first-generation no college, 20% first-generation some college, and 74% not first-generation), Geomarkets that disproportionately larger proportions of first-generation college students whose parents did not attend college include South Bay (11%), South and South Central Los Angeles (21%), Long Beach (7%), and Covina and West Covina (9%). The West Los Angeles and West Beach (26%), Hollywood & Wilshire (36%), East Los Angeles (33%), Long Beach (40%), and Covina and West Covina (27%) Geomarkets also contributed larger proportions of Black first-generation college students whose parents attended college but did not complete their their degree.

Finally, Hispanic prospects in this middle range PSAT order were on average 25% first-generation college students whose parents did not attend college, 25% first-generation college students whose parents attended some college but did not complete their degrees, and 50% were not first-generation college students. The Hollywood & Wilshire (30%), East Los Angeles (39%), South and South Central Los Angeles (54%), and Riverside, San Bernardino, and Ontario (31%) contributed disproportionate shares of Hispanic first-generation college students whose parents did not attend college. The East Los Angeles (33%), South and South Central Los Angeles (28%), Long Beach (30%), and Riverside, San Bernardino, and Ontario (27%) also contributed larger Hispanic first-generation college students whose parents attended college but did not complete their their degree.

\[NOTE SURE YOU NEED THIS PARAGRAPH\]Despite Los Angeles's multi-nuclei urban structure contributing to a more decentralized Geomarket landscape with racial/ethnic diversity across a range of socioeconomic levels more spatially distributed across the metropolitan area, Geomarkets contributions to student prospect pools reflect racial and socioeconomic stratification patterns similar to other metropolitan cases in the study. Affluent Geomarkets like San Fernando Valley–West and West Los Angeles contribute an overrepresentation of White, Asian, and not-first generation college student prospects, the magnitude of which increases substantially as score ranges increase (see Appendix @fig-la-rq2-lowSAT-race and @fig-la-rq2-highSAT-race). On the other hand, more diverse and historically low-income Geomarkets, like South and South Central Los Angeles and Riverside, San Bernardino, and Ontario, contribute disproportionately more Black, Hispanic, and first-generation college student prospects. Similar to Chicago and Dallas-Fort Worth, Los Angeles further solidifies the pattern of some of the lowest-income Geomarkets consistently playing crucial roles in fostering college access for first-generation Asian (East Los Angeles), Black (South and South Central Los Angeles), and Hispanic college students (South and South Central Los Angeles, East Los Angeles, Hollywood and Wilshire, and Riverside, San Berdardino, and Ontario).

>>>>

In comparison, the Glendale and Pasadena (17% versus 14% for all prospects), East Los Angeles (11% versus 5% for all prospects) and Covina and West Covina (17% versus 11% for all prospects) Geomarkets contributed a disproportionate larger share of Asian prospects relative to their contributions to the overall prospect pool.

@@

<!--# [I'M NOT SURE WHERE YOU ARE GETTING THESE NUMBERS FROM; THIS IS THE WRONG FIGURE I THINK. AND WHEN YOU REVISE, THE LEDES ARE RIVSERSIDE, EAST LA AND, SOUTH CENTRAL (ONLY 2% OF PROSPECTS BUT 10% OF NO-COLLEGE) KS-- weird... overall substantive rankings are correct but my %s are all off]. --> @fig-la-rq2-midSAT-ses presents Geomarket contributions by first-generation college student status, demonstrating the disproportionately larger share of first-generation college students contributed to the overall pool by East Los Angeles, South and South Central Los Angeles, and Riverside, San Bernadino, and Ontario Geomarkets. For students whose parents did not attend college, East Los Angeles (16% versus 5% of all prospects), South and South Central Los Angeles (10% versus 2% of all prospects), and Riverside, San Bernardino, and Ontario (24% versus 17% of all prospects) Geomarkets provided a disproportionately larger contribution of first-generation college students. These three Geomarkets also made disproportionately larger contributions of first-generation college student prospects whose parents attended but did not complete their degree, as well as Hollywood & Wilshire (7% versus 5% of all prospects), Long Beach (7% versus 5% of all prospects), and Covina and West Covina (13% versus 10% of all prospects).

<!--#  [I THINK THE WAY YOU WRITE UP THIS FIGURE SHOULD BE CHANGED MORE SUBSTANTIALLY; I ADD SOME COMMENTS BELOW] --> Lastly, @fig-la-rq2-midSAT-combo presents Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity in the Los Angeles metropolitan area.

For Hispanic prospects, the South and South Central Los Angeles Geomarket stands out as providing the most substantial share of first-generation college students. Across all Hispanic prospects in the middle range PSAT orders, 25% were first-generation college students whose parents did not attend college, 25% were first-generation college students whose parents attended some college but did not complete their degrees, and 50% were not first-generation college students. In comparison, the South and South Central Los Angeles Geomarket contributed much larger disproportionate share of Hispanic first-generation college students whose parents did not attend college (54%), with the Hollywood & Wilshire (30%), East Los Angeles (39%), and Riverside, San Bernardino, and Ontario (31%) also contributing larger but more modest disproportionate shares. The East Los Angeles (33%), South and South Central Los Angeles (28%), Long Beach (30%), and Riverside, San Bernardino, and Ontario (27%) also contributed larger Hispanic first-generation college students whose parents attended college but did not complete their their degree.

<!--# [TAKE A TIP FROM THE CRIT QUANT FOLKS; YOU DON'T ALWAYS NEED TO START WITH WHITE; REVISE THIS TEXT TO START WITH A PARAGRAPH ON HISPANIC AND A PARAGRAPH ON ASIAN -- OR VICE VERSA; THOSE TWO ARE THE LEAD; AND THEN CAN TALK ABOUT WHITE AND BLACK. ]Among all White prospects (n=2,511) whose contact information was purchased in this middle PSAT score range, approximately 3% were first-generation college students whose parents did not attend college, 11% were first-generation college students whose parents attended some college but did not complete their degree, and 86% were not first-generation college students. @fig-la-rq2-midSAT-combo shows all Geomarkets contribute a nearly proportionate share of prospects by first-generation college students (across both no college and some college parents) relative to the overall pool. The South and South Central Los Angeles Geomarket stands out as one exception contributing a relatively larger share of White first-generation college students whose parents did not attend college (17%) and whose parent attended some college but did not complete their degree (33%) [DON'T MENTION THIS; THIS IS 6 STUDENTS!]. Other Geomarket that contributed signifcantly larger shares of White first-generation prospects whose parents did not complete their degree include Long Beach (17%), and Riverside, San Bernardino, and Ontario (19%).-->

<!--# [THE LEAD IS EAST LA WHICH HAS 251 ASIAN STUDENTS AND IS WAY DIFFERENT THAN ANY OTHE IN FIRSTGEN; AND THEN WILSHIRE IS SORTA CLOSE]--> Among all Asian prospects whose contact information was purchased, approximately 11% were first-generation college students whose parents did not attend college, 21% were first-generation college students whose parents attended some college but did not complete their degree, and 68% were not first-generation college students. @fig-la-rq2-midSAT-combo highlights the East Los Angeles Geomarket as a key contributor of Asian prospects (n = 251), a significant proportion of whom are first-generation college students (35%) whose parents did not attend college. The Hollywood & Wilshire Geomarket also contributed a modest number of Asian first-generation college students (some college) at a disproportionately larger rate (35%).

While the pool of Black prospects is relatively small (n=186) in @fig-la-rq2-midSAT-combo , some patterns emerge by first-generation status. Overall, approximately 5% of Black prospects were first-generation college students-no college, 20% were first-generation college students-some college, and 74% not first-generation). The South and South Central Los Angeles Geomarket stands out for contributing a disproportionately high share (21%) of Black first-generation college students whose parents did not attend college, despite representing a relatively smaller overall number. Additionally, the Long Beach Geomarket is notable for its higher proportion (40%) of Black first-generation students whose parents attended college but did not complete a degree.

<!--#  [NOTE SURE YOU NEED THIS PARAGRAPH]Despite Los Angeles's multi-nuclei urban structure contributing to a more decentralized Geomarket landscape with racial/ethnic diversity across a range of socioeconomic levels more spatially distributed across the metropolitan area, Geomarkets contributions to student prospect pools reflect racial and socioeconomic stratification patterns similar to other metropolitan cases in the study. Affluent Geomarkets like San Fernando Valley–West and West Los Angeles contribute an overrepresentation of White, Asian, and not-first generation college student prospects, the magnitude of which increases substantially as score ranges increase (see Appendix @fig-la-rq2-lowSAT-race and @fig-la-rq2-highSAT-race). On the other hand, more diverse and historically low-income Geomarkets, like South and South Central Los Angeles and Riverside, San Bernardino, and Ontario, contribute disproportionately more Black, Hispanic, and first-generation college student prospects. Similar to Chicago and Dallas-Fort Worth, Los Angeles further solidifies the pattern of some of the lowest-income Geomarkets consistently playing crucial roles in fostering college access for first-generation Asian (East Los Angeles), Black (South and South Central Los Angeles), and Hispanic college students (South and South Central Los Angeles, East Los Angeles, Hollywood and Wilshire, and Riverside, San Berdardino, and Ontario).-->
